Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-08 paper/analysis.tex

\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
\begin{equation}
    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$. Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as

@@

%The idea behind the proof is to bound  to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality. We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$: 
    \begin{equation}
        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
    \end{equation}
%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
for a Gaussian random vector $\vct{\psi}$.
%
%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. By \cite[Theorem 2.3]{Boucheron2013} this implies for every $k \in \mathbb{N}$ that
    \begin{align*}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{align*}
%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
%     \begin{equation}
%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
%     \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.

@@

%\begin{remark}
%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
%    \begin{equation}
%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
%    \end{equation}
%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
%\end{remark}

@@

    First, we can rewrite
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{(\ell - 1)k + 1}, \dots, \sigma_{\ell k}}_{\leq \sigma_{(\ell - 1)k + 1}}}^{\geq 0}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{(i-1)k + 1}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{(i-2)k + j}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 = \lVert \mtx{\Sigma} \rVert _2^2 + \sum_{i=2}^{\ell} \sigma_{(i-1)k + 1}^2  \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{n=1}^{(\ell - 1)k} \sigma_n^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{(i-1)k + j}^2}_{\leq \sigma_{(i-1)k + 1}^2} \omega_{(i-1)k + j}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{($\sigma_{(i-1)k + j}^2 \leq \sigma_{(i-1)k + 1}^2$, $\forall i, j$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{(triangle inequality)} \notag \\
        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 && \text{(\reflem{lem:gamma})} \notag \\
        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
    \end{align}

    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
    %\begin{align}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
    %\end{align}
    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively. Hence,
    %\begin{equation}
    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
    %    \leq e^2 \sqrt{\frac{3}{2k}}.
    %    \label{equ:OSE-moment-bound-second}
    %\end{equation}

@@

    We can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
    \begin{equation}
        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.

    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by

@@

    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}

@@

    As opposed to \cite[lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$. Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives

@@

    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get 

@@

    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
    \begin{align}
        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{align}
    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
    \end{equation}
    from which follows the first statement.

@@

Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.

>>>>

\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}

\begin{equation}
    \Nystr{\mtx{\Omega}}{\mtx{B}}(t) = (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Then we can estimate the trace as $\Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t))$. Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may rewrite this estimator as

@@

%The idea behind the proof is to bound  to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality. We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$: 
    \begin{equation}
        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
    \end{equation}
%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
for a Gaussian random vector $\vct{\psi}$.
%
%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. By \cite[Theorem 2.3]{Boucheron2013} this implies for every $k \in \mathbb{N}$ that
    \begin{align*}

        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{align*}
%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
%     \begin{equation}
%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
%     \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.

@@

%\begin{remark}
%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
%    \begin{equation}
%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
%    \end{equation}
%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[Remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
%\end{remark}

@@

    First, we can rewrite
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right]
        = \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right].
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \overbrace{\underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}}^{\geq \sigma_{k+1}}, \overbrace{\underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}}^{\geq \sigma_{2k+1}}, \dots, \overbrace{\underbrace{\sigma_{(\ell - 1)k + 1}, \dots, \sigma_{\ell k}}_{\leq \sigma_{(\ell - 1)k + 1}}}^{\geq 0}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{(i-1)k + 1}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{(i-2)k + j}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 = \lVert \mtx{\Sigma} \rVert _2^2 + \sum_{i=2}^{\ell} \sigma_{(i-1)k + 1}^2  \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{n=1}^{(\ell - 1)k} \sigma_n^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\Bigg[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{(i-1)k + j}^2}_{\leq \sigma_{(i-1)k + 1}^2} \omega_{(i-1)k + j}^2 \Bigg] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{($\sigma_{(i-1)k + j}^2 \leq \sigma_{(i-1)k + 1}^2$, $\forall i, j$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 ~ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{(i-1)k + j}^2 \right] && \text{(triangle inequality)} \notag \\
        %&\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k}{2} + p)}{\Gamma(\frac{k}{2})} \right)^{\sfrac{1}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq (k + p) \sum_{i=1}^{\ell} \sigma_{(i-1)k + 1}^2 && \text{(\reflem{lem:gamma})} \notag \\
        &\leq (k + p) \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right). && \text{(using \refequ{equ:singular-value-group-bound})}
    \end{align}

    %The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}
    %\begin{align}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &= \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}} \notag \\
    %    &\leq \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{1}{\Gamma(k - r + 2)} \right)^{\frac{1}{k - r + 1}} \left( \frac{k + r}{2} \right) \notag \\
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right)
    %\end{align}
    %With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively. Hence,
    %\begin{equation}
    %    \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
    %    \leq e^2 \sqrt{\frac{3}{2k}}.
    %    \label{equ:OSE-moment-bound-second}
    %\end{equation}

@@

    We can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
    \begin{equation}
        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.

    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by

@@

    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[Lemma B.3]{tropp-2023-randomized-algorithms}

@@

    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$. Inserting \refequ{equ:spectral-norm-bound-applied} and  \refequ{equ:pinv-spectral-norm-bound} in \refequ{equ:nystrom-proof-processed-tail} with $k = n_{\mtx{\Omega}} / 2$ gives

@@

    We can identify $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 = \lambda_{k+1}$ and $\lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F^2 = \Trace(\mtx{\Lambda}_2)$, and just as in the proof of \cite[Lemma 3.1]{meyer-2021-hutch-optimal} we bound $\lambda_{k+1} \leq \Trace(\mtx{B})/k$ and $\Trace(\mtx{\Lambda}_2) \leq \Trace(\mtx{B})$ -- remembering $k = n_{\mtx{\Omega}}/2$ -- to get 

@@

    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
    \begin{align}
        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
        &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
        &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{align}
    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
    \end{equation}
    from which follows the first statement.

@@

Comparing \refthm{thm:nystrom-pp} with its equivalent for constant matrices \cite[Theorem 3.4]{persson-2022-improved-variants}, we notice a less favorable scaling of the required number of random vectors $n_{\mtx{\Omega}}$ and $n_{\mtx{\Psi}}$ in terms of $\log(\delta^{-1})$.
